A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-09 podc_paper/correctness-2.tex
minor edits

\begin{lemma}\label{RBTupdates}
Suppose a step of the algorithm changes $v.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some $m$.
\end{lemma}
\begin{proof}
The RBT of a node is updated only at line \ref{appendLeafGC} or \ref{casGC}.

If line \ref{appendLeafGC} of an \op{Append} operation modifies $v.\fld{blocks}$,
then $v$ is a leaf node, and no other process ever modifies $v.\fld{blocks}$.
$T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
$B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
Either way, $B.\fld{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
So the claim is satisfied.

If line \ref{casGC} of a \op{Refresh} modifies $v.\fld{blocks}$, then $v$ is an internal node.
After reading $T$ from $v.\fld{blocks}$ at line \ref{refreshReadTGC},
then creates the  block \var{new},
and then calls \op{AddBlock}($T,new$) to create $T'$.
Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
So the claim is satisfied.
\end{proof}

Since each RBT starts with a single block with \fld{index} 0, the following is an easy consequence of \Cref{RBTupdates}.
\begin{corollary}\label{nonEmptyIncreasing}
The RBT stored in each node $v$ is never empty and always stores a set of blocks with consecutive indices.
Moreover, its maximum \fld{index} can only increase over time.
\end{corollary}

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $v.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in $v$'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
each time a new block appears in $v$'s RBT, its \fld{index} is greater than any block
that has appeared in $v$'s RBT earlier.
Thus, $v.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if that block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

@@

%\here{If next obs is only used once, can fold it into the proof that needs it}
%\begin{observation}
%\label{edirGC} 
%If $B$ is a block in the RBT of an internal node $v$, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $v.\var{dir}$.
%\end{observation}
%\begin{proof}
%The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
%on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
%\end{proof}

We next prove the analogue of Lemma \ref{lem::headProgress}.
\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $h>0$ and a block with index $h$ has been inserted into $v.\fld{blocks}$ then 
 $v.\fld{blocks}[h-1].\fld{end\sub{left}} \leq v.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $v.\fld{blocks}[h-1].\fld{end\sub{right}} \leq v.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into $v$'s RBT by the CAS at line \ref{casGC} 
that changed the tree from $T$ to $T'$.
Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $v.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(v,b)$ that created $B$ started.
By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

@@

Next, we prove an analogue of \Cref{successfulRefresh}.
We say a node $v$ \emph{contains} an operation
if some block containing the operation has previously appeared in $v$'s RBT (even if 
the block has been removed from the RBT by a subsequent \op{Split}).

\begin{customlemma}{\ref{successfulRefresh}$'$}
\label{successfulRefreshGC}
Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
In the configuration after that CAS (or termination, respectively), $v$ contains all operations that $v$'s children contained 
when $R$ executed line~\ref{refreshReadTGC}.
\end{customlemma}
\begin{proof}
Suppose $v$'s child (without loss of generality, $v.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{refreshReadTGC}.
Let $i$ be the index of the block containing $op$ that was in $v$'s RBT before 
$R$ executed line \ref{refreshReadTGC}.
We consider two cases.

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
installs $B'$ in $v.\fld{blocks}$.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $v.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in $v$ contains $B$ as a subblock
and therefore $v$ contains $op$.

Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOPGC}.
Let $h$ be the maximum \fld{index} in $T$ plus 1.
By reasoning identical to the last paragraph of \Cref{successfulRefresh}'s proof,
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
that the blocks $v.\fld{left.blocks}[v.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$v.\fld{right.blocks}[v.\fld{blocks}[h-1].\eright+1..new.\eright]$
contain no operations.
By \Cref{blockNotEmpty}, each block contains at least one operation, so
these ranges must be empty, and $v.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
This implies that the block $B$ containing $op$ is a subblock of some block that has appeared
in $v$'s RBT, so $op$ is contained in $v$.
\end{proof}

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}($v$) by the same process.
All operations contained $v$'s children when $R_1$ begins
are contained in $v$ when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from \Cref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $v.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $v.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $v.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
\end{proof}

@@

\begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
Let $B=v.\var{blocks}[b]$.
  If $v.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{customlemma}
\begin{proof}

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
%We say the index $b$ \emph{is marked in node $v$} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a dequeue passes through this block to find the enqueue
%whose argument it should return, or when a dequeue that returns null passes through the block.
%Thus, $b$ gets marked in $v$ when $v.\fld{blocks}[b]$ 
%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
%If the index of block $B$ is marked in $v$, we also say $B$ is marked in $v$.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node $v$'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
$A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

$A$'s call to \op{SplitBlock} determines the index $s$ used to split $v$'s RBT by following 
\eleft\ and \eright\ pointers from the root down to $v$.
So, the block returned is a subblock of block $m$ of the root, unless
at some point along the path of subblocks the subblock has already been removed by a split,
in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.

Next, we bound the number of operations in $v$'s blocks that are retained when $A$ does sets
$T'\assign\op{Split}(T,s)$ operation.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
There are at most $p$ operations in block $m$ of the root by \Cref{blockSizeGC}
and we showed above that there are at most $2q_{max} + 2p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
block is inserted before $A$ sets $v$'s \fld{blocks} to the resulting RBT.
\end{proof}

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
\here{There may be room to improve this bound by a more careful analysis since the bounds so far
didn't use the fact that operations only appear along one path in the tree; in other words
doing a bound on total size of all RBTs at one level of the ordering tree at once, rather than
on each node separately, might reduce it to qlogp + poly(p).  Might also be possible to be
more precise about how it depends on "current" size of queue rather than max size.}

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}.
If a process does GC in a call to \op{AddBlock}($v,T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on $v$}.
We argue that each process $P$ can do an $r$th helping phase on $v$ at most once as follows.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on $v$.
Let $v,T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node $v$ is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes $v.blocks$ from $T$ to another tree.
Either way, by \Cref{RBTupdates}, $v$'s RBT's maximum \fld{index} will be at least $rG$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on $v$ again, the block it passes as the third argument
will have $\index>rG$, so $P$ will not perform an $r$th helping phase on $v$ again.

Each helping phase takes $O(p)$ steps to read the \var{last} array,
$O(p \log p \log(p+q_{max}))$ steps to \op{Help},
followed by $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
Thus, for each integer $r$ and each node $v$, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on $v$.
We can amortize these steps over the operations that appear in 
$v.blocks[(r-1)p^2+1..p^2]$.
By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$, which is
within a constant factor of the steps it performs excluding GC.
\end{proof}

>>>>

\begin{lemma}\label{RBTupdates}
Suppose a step of the algorithm changes $\var{v}.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some $m$.
\end{lemma}
\begin{proof}
The RBT of a node is updated only at line \ref{appendLeafGC} or \ref{casGC}.

If line \ref{appendLeafGC} of an \op{Append} operation modifies $\var{v}.\fld{blocks}$,
then \var{v} is a leaf node, and no other process ever modifies $\var{v}.\fld{blocks}$.
$T'$ was obtained from $T$ by calling \op{AddBlock}($T,B$).
$B$ was created either by the \op{Enqueue} or \op{Dequeue} that called \op{Append}.
Either way, $B.\fld{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $B$ to it.
So the claim is satisfied.

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
then creates the  block \var{new},
and then calls \op{AddBlock}($T,new$) to create $T'$.
Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
So the claim is satisfied.
\end{proof}

Since each RBT starts with a single block with \fld{index} 0, the following is an easy consequence of \Cref{RBTupdates}.
\begin{corollary}\label{nonEmptyIncreasing}
The RBT stored in each node \var{v} is never empty and always stores a set of blocks with consecutive indices.
Moreover, its maximum \fld{index} can only increase over time.
\end{corollary}

Since RBTs are always non-empty, calls to \op{MaxBlock} have well-defined answers.  
Throughout the proof, we use $\var{v}.\blocks[b]$ to refer to the block with \fld{index} $b$ that appeared
in \var{v}'s tree at some time during the execution.
It follows from \Cref{RBTupdates} and \Cref{nonEmptyIncreasing} that 
each time a new block appears in \var{v}'s RBT, its \fld{index} is greater than any block
that has appeared in \var{v}'s RBT earlier.
Thus, $\var{v}.\blocks[b]$ is unique, if it exists.
We also use this notation in the code to indicate that a search of the RBT should be performed for the block
with \fld{index} $b$.
We describe later what should happen if that block is not found by the search because it has been removed
by a GC phase.
\here{remember to come back to this point}

@@

%\here{If next obs is only used once, can fold it into the proof that needs it}
%\begin{observation}
%\label{edirGC} 
%If $B$ is a block in the RBT of an internal node \var{v}, then for $\var{dir} \in \{\fld{left},\fld{right}\}$,
%$B.\edir$ was read from the \fld{index} field of some node in the RBT of $\var{v}.\var{dir}$.
%\end{observation}
%\begin{proof}
%The \eleft\ and \eright\ fields are set by reading the \fld{index} field of a node in the appropriate child
%on line \ref{createEndLeftGC} and \ref{createEndRightGC}.
%\end{proof}

We next prove the analogue of Lemma \ref{lem::headProgress}.
\begin{customlemma}{\ref{lem::headProgress}$'$}\label{lem::headProgressGC}
 If $h>0$ and a block with index $h$ has been inserted into $\var{v}.\fld{blocks}$ then 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{left}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{left}}$ and 
 $\var{v}.\fld{blocks}[h-1].\fld{end\sub{right}} \leq \var{v}.\fld{blocks}[h].\fld{end\sub{right}}$.
\end{customlemma}
\begin{proof}
The block $B$ with index $h$ was installed into \var{v}'s RBT by the CAS at line \ref{casGC} 
that changed the tree from $T$ to $T'$.
Before this CAS, line \ref{refreshReadTGC} read the tree $T$ from $\var{v}.blocks$,
line \ref{refreshReadMax} found a block $B'$  with \fld{index} $h-1$ in $T$,
and then line \ref{invokeCreateBlockGC} created the block $B$ with $\fld{index}=h$.
Since $B'$ was already in $T$ before $B$ was created, the
\op{CreateBlock}$(v,b-1)$ that created $B'$ terminated before the 
\op{CreateBlock}$(v,b)$ that created $B$ started.
By \Cref{nonEmptyIncreasing}, the value that line \ref{createEndLeftGC} of \op{CreateBlock}$(v,b-1)$ 
stores in $B'.\eleft$ is less than or equal to the value that line \ref{createEndLeftGC} 
of \op{CreateBlock}$(v,b)$ stores in $B.\eleft$.  
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRightGC} satisfy the claim.
\end{proof}

@@

Next, we prove an analogue of \Cref{successfulRefresh}.
We say a node \var{v} \emph{contains} an operation
if some block containing the operation has previously appeared in \var{v}'s RBT (even if 
the block has been removed from the RBT by a subsequent \op{Split}).

\begin{customlemma}{\ref{successfulRefresh}$'$}
\label{successfulRefreshGC}
Let $R$ be a call to \op{Refresh}(\var{v}) that performs a successful \op{CAS} on line \ref{casGC} (or terminates at line \ref{addOPGC}).
In the configuration after that CAS (or termination, respectively), \var{v} contains all operations that \var{v}'s children contained 
when $R$ executed line~\ref{refreshReadTGC}.
\end{customlemma}
\begin{proof}
Suppose \var{v}'s child (without loss of generality, $\var{v}.\fld{left}$) contained an operation $op$ 
when $R$ executed line \ref{refreshReadTGC}.
Let $i$ be the index of the block containing $op$ that was in \var{v}'s RBT before 
$R$ executed line \ref{refreshReadTGC}.
We consider two cases.

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
installs $B'$ in $\var{v}.\fld{blocks}$.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
and therefore \var{v} contains $op$.

Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOPGC}.
Let $h$ be the maximum \fld{index} in $T$ plus 1.
By reasoning identical to the last paragraph of \Cref{successfulRefresh}'s proof,
it follows from the fact that $\var{num\sub{enq}}+\var{num\sub{deq}}=0$ at line \ref{testEmptyGC}
that the blocks $\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..new.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..new.\eright]$
contain no operations.
By \Cref{blockNotEmpty}, each block contains at least one operation, so
these ranges must be empty, and $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{new}.\eleft\geq i$.
This implies that the block $B$ containing $op$ is a subblock of some block that has appeared
in \var{v}'s RBT, so $op$ is contained in \var{v}.
\end{proof}

@@

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from \Cref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from \Cref{successfulRefreshGC} applied to $R_3$.
\end{proof}

@@

\begin{customlemma}{$\ref{superRelation}'$}\label{superRelationGC}
Let $B=\var{v}.\fld{blocks}[b]$.
  If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{customlemma}
\begin{proof}

@@

We first bound the size of RBTs.  Let $q_{max}$ be the maximum size of the queue at any time during the sequential execution given by the linearization $L$.
Recall that GC is done on a node every $G$ times its RBT is updated, and we chose $G$ to be $p^2\ceil{\log p}$.
Part of the proof of the following lemma is similar to the proof of \Cref{dSearchTime}.
%We say the index $b$ \emph{is marked in node \var{v}} if some process has tested
%whether $b>\var{v}.\fld{last}[*]$ at line \ref{marktest1} or \ref{marktest2}.
%Intuitively, a block's index value is marked 
%when a dequeue passes through this block to find the enqueue
%whose argument it should return, or when a dequeue that returns null passes through the block.
%Thus, $b$ gets marked in \var{v} when $\var{v}.\fld{blocks}[b]$ 
%contains either an enqueue whose value is being dequeued or a \nl\ dequeue.
%If the index of block $B$ is marked in \var{v}, we also say $B$ is marked in \var{v}.

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
$A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

$A$'s call to \op{SplitBlock} determines the index $s$ used to split \var{v}'s RBT by following 
\eleft\ and \eright\ pointers from the root down to \var{v}.
So, the block returned is a subblock of block $m$ of the root, unless
at some point along the path of subblocks the subblock has already been removed by a split,
in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$ does sets
$T'\assign\op{Split}(T,s)$ operation.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
There are at most $p$ operations in block $m$ of the root by \Cref{blockSizeGC}
and we showed above that there are at most $2q_{max} + 2p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
Since each block is non-empty by \Cref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
\here{For later version:  There may be room to improve this bound by a more careful analysis since the bounds so far
didn't use the fact that operations only appear along one path in the tree; in other words
doing a bound on total size of all RBTs at one level of the ordering tree at once, rather than
on each node separately, might reduce it to qlogp + poly(p).  Might also be possible to be
more precise about how it depends on "current" size of queue rather than max size.}

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.
We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once as follows.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by \Cref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $rG$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\index>rG$, so $P$ will not perform an $r$th helping phase on \var{v} again.

Each helping phase takes $O(p)$ steps to read the \var{last} array,
$O(p \log p \log(p+q_{max}))$ steps to \op{Help},
followed by $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on \var{v}.
We can amortize these steps over the operations that appear in 
$\var{v}.\fld{blocks}[(r-1)p^2+1..p^2]$.
By \Cref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$, which is
within a constant factor of the steps it performs excluding GC.
\end{proof}

### 2023-01-09 podc_paper/introduction.tex
minor edits

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21} \here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
\here{$O(?)$ space} and \here{$O(?)$ steps per operation}.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

>>>>

There has been a great deal of research in the past several decades on the design of linearizable, lock-free queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management (e.g., \cite{BBFRSW21}\here{try to find an older, more canonical reference}),
packet processing \cite{DPDK},
and sharing resources or tasks.
The lock-free MS-queue of Michael and Scott \cite{MS98} is a classic shared queue implementation.
It uses a singly-linked list with pointers to the front and back nodes.
To dequeue or enqueue an element, the front or back pointer is updated by a 
compare-and-swap (CAS) instruction.
If this CAS fails, the operation must retry.
In the worst case, this means that each successful CAS may cause all other processes to
fail and retry, leading to an amortized step complexity of $\Omega(p)$ per operation.
(To measure amortized step complexity of a lock-free implementation, we consider all possible finite executions
and divide the number of steps in the execution by the number of operations initated in the execution.)
Numerous papers have suggested modifications to the MS-queue \cite{DBLP:conf/opodis/HoffmanSS07,DBLP:conf/podc/KoganH14,DBLP:conf/ppopp/KoganP11,DBLP:journals/dc/Ladan-MozesS08,MKLLP22,DBLP:conf/spaa/MoirNSS05,RC17}, but 
all still have $\Omega(p)$ amortized step complexity as a result of
contention on the front and back of the queue.
Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} called this the \emph{CAS retry problem}.
The same problem occurs in array-based implementations of queues \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01,DBLP:conf/opodis/GidenstamST10}.
Solutions that tried to sidestep this problem using fetch\&increment \cite{DBLP:conf/ppopp/MorrisonA13,DBLP:conf/ppopp/YangM16,Nik19,10.1145/3490148.3538572}
rely on slower mechanisms to handle worst-case executions and still have $\Omega(p)$ step complexity.

@@

We exploit this loophole.  We show that it is, in fact, possible to implement a linearizable, wait-free queue
whose step complexity does not depend linearly on $p$.
Our implementation is the first with step complexity that is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
For ease of presentation, we first give an unbounded-space construction that uses
$O(\log^2 p + \log q)$ steps per operation.
Then, we show how to modify the implementation to use 
bounded space while still having amortized step complexity $O(\log^2p + \log q)$ per operation.
Both implementations use single-word CAS and reasonably-sized words.
Moreover, they use $O(\log p)$ CAS instructions per operation in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
For the space-bounded implementation, we unlink unneeded memory from our data structure.
We do not address the orthogonal problem of reclaiming unlinked memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.
